Import libraries
```{r}
library(readr)
```


USING THE CRITICS DATASET

Load the critics dataset
```{r}
critics <- read.csv('critics.csv', sep=',', dec=',')
critics <- as.data.frame(critics)
colnames(critics) <- gsub('.', ' ', colnames(critics), fixed=T)
critics
```

Get the top 5 movies, ordered by the mean of their ratings
```{r}
# Calculate the mean of each column(movie), except for the first one (the User column)
means1 <- colMeans(critics[,-1], na.rm=T)

# Store the column names of the original critics dataframe, except for the first column (User)
column <- colnames(critics)
columns <- column[2:21]

# Create a new dataframe called means2 and store the mean rating for each movie in that dataframe
means2 <- data.frame(movie_name=columns, rating=means1, row.names=NULL)

# Order this previous dataframe and store the top 5 movies (by rating) in a new dataframe called means4
means3 <- means2[order(means2$rating,decreasing=T), ]
means4 <- means3[1:5,]

# Print the top 5 movies and their ratings
paste(means4$rating, means4$movie_name, sep=',')
```

Get the top 5 movies, ordered by their ratings
```{r}
# Calculate, for each movie, the number of rows that are not NA (the number of ratings given for each movie)
count <- sapply(critics[,-1], function(x) {sum(!is.na(x))})

# Calculate, for each movie, the number of ratings that are greater than or equal to 4
greater <- sapply(critics[,-1], function(x) {length(which(x>=4))})

# Calculate, for each movie, the percentage of ratings that are greater than or equal to 4 and store the output in a dataframe
ratings <- greater / count
ratings2 <- data.frame(ratings)
top <- data.frame(movies=rownames(ratings2),ratings,row.names=NULL)

# Order the movies by rating distribution and store the top 5 in a new dataframe called top5
top2 <- top[order(top$ratings,decreasing=T),]
top5 <- top2[1:5,]

# Print the top 5 movies by rating, along with their rating
paste(top5$ratings,top5$movies,sep=',')
```

Get the top 5 movies, ordered by the number of ratings they received
```{r}
# Calculate, for each movie, the number of rows that are not NA (the number of ratings given for each movie) and store the output in a 
# dataframe
counts <- sapply(critics[,-1],function(x) {sum(!is.na(x))})
counts2 <- data.frame(counts)
counts3 <- data.frame(movies=rownames(counts2),numb_ratings=counts,row.names=NULL)

# Order the movies by quantity of ratings and store the top 5 in a new dataframe called ordered2
ordered <- counts3[order(counts3$numb_ratings,decreasing=T),]
ordered2 <- ordered[1:5,]

# Print the top 5 movies by quantity of ratings, along with their ratings
paste(ordered2$numb_ratings,ordered2$movies,sep=',')
```

Get the top 5 movies recommended to users who also watched The Wizard of Oz
```{r}
# Remove the first column of the critics dataframe. It contains user names
critics2 <- critics[,-1]

# Subset the previous dataframe to contain the rows where there are no NA values for the Star Wars IV   A New Hope movie. Since we're comparing the 
# percentage of other movie raters who also rated that movie, we need to keep the rows of the users who rate that movie, hence no NA
critics3 <- critics2[which(!is.na(critics2$`Star Wars IV   A New Hope`)), ]

# Calculate the number of users who rated each movie, divided by the number of users who rated Star Wars IV   A New Hope. The result, which is the 
# percentage of other movie raters who also rated Star Wars IV   A New Hope is stored in a dataframe
percents <- sapply(critics3, function(x){sum(!is.na(x))/nrow(critics3)})  
percents2 <- data.frame(percents)
percents3 <- data.frame(movies=rownames(percents2),occurences=percents,row.names=NULL)

# Order the movies based on the highest percentage 
p_order <- percents3[order(percents3$occurences,decreasing=T),]

# Store the top 5 in a new dataframe called top5. We start the index at 2 because our reference movie, Star Wars IV   A New Hope, has the highest 
# result of 1 because it's compared to itself
p_order1 <- p_order[2:6,]

# Print the top 5 movies by percentage of other users who also rated Star Wars IV   A New Hope
paste(p_order1$occurences,p_order1$movies,sep=',')
```

Get the top 5 movie recommendations for users who also liked Babe
```{r}
# Subset the original critics dataframe to only include the rows of users who rated the movie Babe (no NA values)
critics4 <- critics[!is.na(critics$Babe),c('User','Babe')]

# Subset the previous dataframe to keep the rows of users who gave either 4 or 5 stars to Babe, meaning that they liked the movie
critics5 <- critics4[(critics4$Babe==4|critics4$Babe==5),]

# Subset the previous dataframe to remove the Babe column because it is our reference movie
critics6 <- critics[,-ncol(critics)] 

# Subset the previous dataframe to keep the users who rated Babe with 4 or 5 stars, but keeping all the movies, except for Babe
critics7 <- critics6[which(critics$User %in% critics5$User),]

# For these users, calculate the average rating they gave to other movies, and store the result in a dataframe
mean1 <- colMeans(critics7[,-1], na.rm=T)
mean2 <- data.frame(movie=colnames(critics6[,-1]),rating=mean1,row.names=NULL)

# Order the dataframe by ratings and select the top 5
mean3 <- mean2[order(mean2$rating,decreasing=T),]
mean4 <- mean3[1:5,]

# Print the top 5 movies for people who gave 4 or 5 stars to Babe
paste(mean4$rating,mean4$movie,sep=',')
```


USING ANOTHER DATASET, MOVIELENS, WHICH HAS A DIFFERENT STORAGE LAYOUT THAN THE CRITICS DATASET WE HAVE BEEN WORKING WITH PREVIOUSLY

Load the MovieLens dataset
```{r}
movies <- read.csv("movies.csv", sep=',')
ratings <- read.csv("ratings.csv", sep=',')
head(movies)
head(ratings)
```

Merge both datasets into a new one and remove the movieId, genres, and timestamp variables
```{r}
mr0 <- merge(movies, ratings, by.x='movieId')
mr <- mr0[,-c(1,3,6)]
```

Filter the dataframe and keep the rows/movies that are rated by at least 100 users
```{r}
agg <- aggregate(mr$rating, by=list(mr$title), FUN=function(x) {length(x)})
agg1 <- subset(agg, x>100)
agg2 <- agg1[, -2]
mr1 <- mr[mr$title %in% agg2,]
head(mr1)
```

Get the top 5 movies ordered by the mean of their ratings
```{r}
# Use the aggregate function to get the mean rating of each movie, grouping the result by movie
mean_r <- aggregate(mr1$rating, by=list(mr1$title), FUN=function(x) {sum(x)/length(x)})

# Convert the output to a data frame and ordering the ratings and keeping the top 5 results
mean_r1 <- data.frame(movies=mean_r$Group.1, ratings=mean_r$x, row.names=NULL)
mean_r2 <- mean_r1[order(mean_r1$ratings, decreasing=T), ]
mean_r3 <- mean_r2[1:5,]

# Print the top 5 movies by movie ratings
paste(mean_r3$ratings, mean_r3$movies, sep=',')
```

Get the top 5 movies, ordered by their ratings 
```{r}
# For each movie, calculate the percentage of ratings that are 4 stars or grater
percent_r <- aggregate(mr1$rating, by=list(mr1$title), FUN=function(x) {(length(which(x>=4)))/(sum(!is.na(x)))})

# Rename the columns of this new data frame
colnames(percent_r) <- c('movie', 'percent >= 4stars')

# Order the movies by rating distribution 
percent_r1 <- percent_r[order(percent_r$`percent >= 4stars`, decreasing=T),]
percent_r2 <- percent_r1[1:5,]

# Print the top 5 movies by rating distribution
paste(percent_r2$`percent >= 4stars`, percent_r2$movie, sep=',')
```

Get the top 5 movies, ordered by the number of ratings they received
```{r}
# For each movie, calculate the number of rows that are not NA
counts_r <- aggregate(mr1$rating, by=list(mr1$title), FUN=function(x) {sum(!is.na(x))})
colnames(counts_r) <- c('movie', 'number_of_ratings')

# Order the movies by quantity of ratings
order_r <- counts_r[order(counts_r$number_of_ratings, decreasing=T), ]
order_r1 <- order_r[1:5,]

# Print the top 5 movies by quantity of ratings
paste(order_r1$number_of_ratings, order_r1$movie, sep=',')
```

Get the top 5 movies recommended to users who also watched Toy Story
```{r}
# Subset the initial dataframe mr1 to include the users who rated Toy Story
mr2 <- mr1[mr1$title=='Toy Story (1995)',]
ts <- mr1[which(mr1$userId %in% mr2$userId), ]

# Calculate the number of users who rated each movie, divided by the number of users who rated Toy Story. The result is the percentage of 
# other movie raters who also rated Toy Story
numb_users <- length(unique(mr2$userId))
ts1 <- aggregate(ts$userId, by=list(ts$title), FUN=function(x) {(length(unique(x)))/numb_users})
colnames(ts1) <- c('movie', 'percentage')

# Order the movies based on the highest percentage 
ts2 <- ts1[order(ts1$percentage, decreasing=T), ]

# Store the top 5 in a new dataframe called ts3. We start the index at 2 because our reference movie, Toy Story, has the highest result of
# 1 because it's compared to itself
ts3 <- ts2[2:6,]

# Print the top 5 movies that people who watched Toy Story also watched
paste(ts3$percentage, ts3$movie, sep=',')
```

Get the top 5 movie recommendations for users who also liked Toy Story
```{r}
# Subset our mr1 dataframe to keep the rows for the movie Toy Story, meaning that we want to keep the users who have rated this movie
mr2 <- mr1[mr1$title=='Toy Story (1995)',]

# Subset the previous dataframe to keep the rows of users who gave either 4 or 5 stars to Toy Story
mr3 <- mr2[(mr2$rating==4|mr2$rating==5),]

# Subset the initial dataframe mr1 to remove the rows for the movie Toy Story because it is our reference movie
mr4 <- mr1[!(mr1$title=='Toy Story (1995)'), ]

# Subset the previous dataframe to keep the users who rated Toy Story with 4 or 5 stars, but keeping all the movies, except for Toy Story
mr5 <- mr4[which(mr4$userId %in% mr3$userId), ]

# For these users, calculate the average rating they gave to other movies
avg_rating <- aggregate(mr5$rating, by=list(mr5$title), FUN=function(x) {sum(x)/length(x)})
colnames(avg_rating) <- c('movie', 'average_rating')

# Order the dataframe by ratings and select the top 5
avg_rating1 <- avg_rating[order(avg_rating$average_rating, decreasing=T), ]
avg_rating2 <- avg_rating1[1:5, ]

# Print the top 5 movies that people who watched Toy Story also liked
paste(avg_rating2$average_rating, avg_rating2$movie, sep=',')
```